---
title: "Hypoxia and mTOR Inhibition Reshape the Transcriptome of Normal and Triple Negative Breast Cancer Cells"
subtitle: "A multifactor RNA seq differential expression analysis (GSE104193)"
author: "Freya Montgomerie"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                       fig.width = 8, fig.height = 6, cache = TRUE)
library(tidyverse)
library(DESeq2)
library(GEOquery)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
library(msigdbr)
library(enrichplot)
library(knitr)
```

# 1. Background and Motivation

Hypoxia is a defining feature of the tumour microenvironment. Cancer cells frequently
alter their transcriptional and translational programmes to survive under low oxygen
conditions. The mTOR pathway is a central regulator of protein synthesis and is closely
linked to oxygen and nutrient sensing. This makes mTOR inhibitors, such as PP242, of
interest both as research tools and as potential cancer therapeutics.

This project analyses a public RNA seq dataset ([GSE104193](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE104193)),
comparing a non malignant breast epithelial cell line (MCF10A) with a triple negative
breast cancer cell line (MDA-MB-231). Cells were exposed to normoxia or hypoxia, with
and without the mTORC1/2 inhibitor PP242, and both total and polysome bound mRNA were
profiled. This gives a factorial design (2 cell lines x 2 oxygen levels x 2 drug
conditions x 2 mRNA fractions, 2 biological replicates each, 32 samples total), which
was used to ask:

1. How does hypoxia change gene expression, and does this differ between the normal and
   cancer cell lines?
2. Is hypoxia's effect on transcription also seen proportionally in the polysome bound
   (translating) mRNA pool, or is there evidence of hypoxia specific translational
   control?

# 2. Data and Methods

## 2.1 Data retrieval

I retrieved the raw RSEM expected counts and sample metadata directly from GEO using the
`GEOquery` package. Two supplementary count files were provided, split by cell line,
which I merged into a single 60,554 gene by 32 sample count matrix. Sample identity
across the two files could not be assumed from row or column order, so this was resolved
using the explicit `barcode:ch1` metadata field instead. I checked the mapping was
correct by confirming exact agreement between the metadata sample IDs and the count
matrix column names before proceeding.

```{r data-retrieval}
gse <- getGEO("GSE104193", GSEMatrix = TRUE)
gse <- gse[[1]]
pheno <- pData(gse)

supp_dir <- "data/raw/GSE104193"
supp_files_expected <- file.path(supp_dir, c(
  "GSE104193_COUNTS_POLYSOME_01_new_pipeline.txt.gz",
  "GSE104193_COUNTS_POLYSOME_02_new_pipeline.txt.gz"
))

if (!all(file.exists(supp_files_expected))) {
  getGEOSuppFiles("GSE104193", baseDir = "data/raw")
} else {
  message("Supplementary files already present locally — skipping re-download.")
}

counts_01 <- read.delim(supp_files_expected[1],
                         header = TRUE, row.names = 1, check.names = FALSE)
counts_02 <- read.delim(supp_files_expected[2],
                         header = TRUE, row.names = 1, check.names = FALSE)
```

## 2.2 Sample metadata construction

```{r metadata}
sample_info <- pheno %>%
  rownames_to_column("geo_accession_row") %>%
  transmute(
    gsm = geo_accession,
    barcode = `barcode:ch1`,
    cell_line = `cell line:ch1`,
    fraction = `fraction:ch1`,
    replicate = `replicate:ch1`,
    treatment_raw = `treatment:ch1`
  ) %>%
  separate(treatment_raw, into = c("oxygen", "drug"), sep = "\\+", fill = "right") %>%
  mutate(
    drug = ifelse(is.na(drug), "Control", drug),
    fraction = recode(fraction, "Total mRNA" = "Total", "Polysomal mRNA" = "Polysomal"),
    replicate = recode(replicate, "Biological replicate 1" = "Rep1", "Biological replicate 2" = "Rep2")
  )

counts_combined <- cbind(counts_01, counts_02)
barcode_to_gsm <- setNames(sample_info$gsm, sample_info$barcode)
colnames(counts_combined) <- barcode_to_gsm[colnames(counts_combined)]
counts_combined <- counts_combined[, sample_info$gsm]

stopifnot(all(colnames(counts_combined) == sample_info$gsm))

coldata <- sample_info
rownames(coldata) <- coldata$gsm
coldata$cell_line <- factor(coldata$cell_line, levels = c("MCF10A", "MDA-MB-231"))
coldata$fraction  <- factor(coldata$fraction, levels = c("Total", "Polysomal"))
coldata$oxygen    <- factor(coldata$oxygen, levels = c("Normoxia", "Hypoxia"))
coldata$drug      <- factor(coldata$drug, levels = c("Control", "PP242"))
coldata$replicate <- factor(coldata$replicate)

kable(head(coldata))
```

## 2.3 Differential expression model

Counts were rounded to integers, as required by DESeq2, because RSEM produces expected
counts rather than integer counts. Genes with fewer than 10 reads in at least 4 samples
were removed before modelling.

I used a multi-factor design:

```
~ cell_line + oxygen + drug + fraction + cell_line:oxygen + oxygen:fraction
```

I did not fit a fully saturated four way factorial model. With only 2 biological
replicates per condition, an unconstrained model would leave very few residual degrees
of freedom for reliable dispersion estimation. Instead, the interaction terms included
were chosen based on the specific questions in Section 1. The `cell_line:oxygen` term
tests whether the hypoxia response differs between the normal and cancer cell lines. The
`oxygen:fraction` term tests whether hypoxia's effect on transcription is also seen
proportionally in the polysome bound (translating) mRNA pool.

```{r deseq2-model}
counts_int <- round(as.matrix(counts_combined))

dds <- DESeqDataSetFromMatrix(
  countData = counts_int,
  colData = coldata,
  design = ~ cell_line + oxygen + drug + fraction + cell_line:oxygen + oxygen:fraction
)
keep <- rowSums(counts(dds) >= 10) >= 4
dds <- dds[keep, ]
dds <- DESeq(dds)

nrow(dds)
resultsNames(dds)
```

Filtering retained `r nrow(dds)` of the original `r nrow(counts_int)` genes for
modelling. Effect sizes were shrunk using apeglm (Zhu, Ibrahim and Love, 2018) before
ranking and visualisation.

# 3. Quality Control

## 3.1 Library sizes

```{r library-sizes, fig.height=5}
library_sizes <- colSums(counts(dds))
barplot(sort(library_sizes), las = 2, cex.names = 0.5,
        main = "Library sizes per sample", ylab = "Total counts")
```

Library sizes ranged from `r round(min(library_sizes)/1e6, 1)`M to
`r round(max(library_sizes)/1e6, 1)`M reads across the 32 samples. This was accounted for
using DESeq2's median of ratios size factor normalisation.

## 3.2 Sample structure (PCA)

```{r pca}
vsd <- vst(dds, blind = TRUE)
pca_data <- plotPCA(vsd, intgroup = c("cell_line", "oxygen", "drug", "fraction"), returnData = TRUE)
percentVar <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(x = PC1, y = PC2, color = cell_line, shape = fraction)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_bw() +
  labs(title = "PCA colored by cell line, shaped by fraction")

ggplot(pca_data, aes(x = PC1, y = PC2, color = oxygen, shape = drug)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_bw() +
  labs(title = "PCA colored by oxygen, shaped by drug")
```

PC1 (`r percentVar[1]`% of variance) separated samples by cell line. This is expected,
as MCF10A (non malignant) and MDA-MB-231 (triple negative cancer) are very different cell
types at the transcriptional level. PC2 (`r percentVar[2]`% of variance) separated
samples by oxygen level within both cell lines. Although this is a small proportion of
the total variance, the separation was consistent across both cell lines, which suggests
a real hypoxia signature. Neither drug nor mRNA fraction produced a visible separation on
the top two PCs. This suggests these effects are smaller, and may be restricted to
specific groups of genes rather than affecting the whole transcriptome.

## 3.3 Dispersion estimates

```{r dispersion}
plotDispEsts(dds)
```

Gene wise dispersion estimates shrank towards the fitted trend, as expected. A small
number of genes showed unusually low dispersion estimates at low mean counts. This is a
known artefact caused by rounding the non integer RSEM expected counts to integers, and
is corrected by DESeq2's shrinkage.

# 4. Results

## 4.1 Hypoxia response in normal (MCF10A) cells

```{r hypoxia-mcf10a}
res_hypoxia_MCF10A <- results(dds, name = "oxygen_Hypoxia_vs_Normoxia")
res_hypoxia_MCF10A_shrunk <- lfcShrink(dds, coef = "oxygen_Hypoxia_vs_Normoxia", type = "apeglm")
summary(res_hypoxia_MCF10A)

gene_info <- data.frame(
  full_id = rownames(dds),
  ensembl_id = sapply(strsplit(rownames(dds), ","), `[`, 1),
  gene_symbol = sapply(strsplit(rownames(dds), ","), `[`, 2),
  biotype = sapply(strsplit(rownames(dds), ","), `[`, 3)
)

volcano_df <- as.data.frame(res_hypoxia_MCF10A_shrunk)
volcano_df$gene_symbol <- gene_info$gene_symbol[match(rownames(volcano_df), gene_info$full_id)]
volcano_df$sig <- ifelse(!is.na(volcano_df$padj) & volcano_df$padj < 0.05 & abs(volcano_df$log2FoldChange) > 1,
                          "Significant", "Not significant")

ggplot(volcano_df, aes(x = log2FoldChange, y = -log10(pvalue), color = sig)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("Significant" = "firebrick", "Not significant" = "grey70")) +
  theme_bw() +
  labs(title = "Hypoxia vs Normoxia (MCF10A, Total mRNA)",
       x = "log2 Fold Change (shrunk)", y = "-log10(p-value)")

top_hypoxia_genes <- volcano_df %>%
  filter(!is.na(padj)) %>%
  arrange(padj) %>%
  dplyr::select(gene_symbol, log2FoldChange, padj) %>%
  head(20)

kable(top_hypoxia_genes, digits = 3)
```

Hypoxia caused widespread transcriptional changes in MCF10A cells. The genes with the
strongest induction included known HIF target genes: CA9 and CA12 (pH regulation),
EGLN3 (part of the oxygen sensing feedback loop), BNIP3L (hypoxia induced mitophagy), and
the glycolysis genes PFKFB4 and PGK1. This shows the pipeline correctly identifies known
hypoxia biology, which supports the validity of the wider analysis.

```{r heatmap, fig.height=8}
top30 <- volcano_df %>% filter(!is.na(padj)) %>% arrange(padj) %>% head(30)
mcf10a_total_samples <- rownames(coldata)[coldata$cell_line == "MCF10A" & coldata$fraction == "Total"]
mat <- assay(vsd)[rownames(top30), mcf10a_total_samples]
mat_scaled <- t(scale(t(mat)))
rownames(mat_scaled) <- top30$gene_symbol

annotation_col <- data.frame(
  Oxygen = coldata[mcf10a_total_samples, "oxygen"],
  Drug = coldata[mcf10a_total_samples, "drug"]
)
rownames(annotation_col) <- mcf10a_total_samples

pheatmap(mat_scaled, annotation_col = annotation_col,
         main = "Top 30 Hypoxia-Responsive Genes (MCF10A, Total mRNA)")
```

## 4.2 Hypoxia response differs between normal and cancer cells

```{r hypoxia-interaction}
res_interaction_cellline_oxygen <- results(dds, name = "cell_lineMDA.MB.231.oxygenHypoxia")
res_interaction_cellline_oxygen_shrunk <- lfcShrink(dds, coef = "cell_lineMDA.MB.231.oxygenHypoxia", type = "apeglm")
res_hypoxia_MDAMB231 <- results(dds, contrast = list(c("oxygen_Hypoxia_vs_Normoxia", "cell_lineMDA.MB.231.oxygenHypoxia")))

summary(res_interaction_cellline_oxygen)
summary(res_hypoxia_MDAMB231)

interaction_df <- as.data.frame(res_interaction_cellline_oxygen_shrunk)
interaction_df$gene_symbol <- gene_info$gene_symbol[match(rownames(interaction_df), gene_info$full_id)]

ggplot(interaction_df, aes(x = log2FoldChange, y = -log10(pvalue),
                            color = !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey70"), guide = "none") +
  theme_bw() +
  labs(title = "Cell line x Oxygen interaction",
       x = "log2 Fold Change (shrunk)", y = "-log10(p-value)")

top_interaction_genes <- interaction_df %>%
  filter(!is.na(padj)) %>%
  arrange(padj) %>%
  dplyr::select(gene_symbol, log2FoldChange, padj) %>%
  head(20)

mdamb231_df <- as.data.frame(res_hypoxia_MDAMB231)
mdamb231_df$gene_symbol <- gene_info$gene_symbol[match(rownames(mdamb231_df), gene_info$full_id)]

comparison_table <- top_interaction_genes %>%
  dplyr::select(gene_symbol, interaction_LFC = log2FoldChange, interaction_padj = padj) %>%
  left_join(volcano_df %>% dplyr::select(gene_symbol, MCF10A_hypoxia_LFC = log2FoldChange), by = "gene_symbol") %>%
  left_join(mdamb231_df %>% dplyr::select(gene_symbol, MDAMB231_hypoxia_LFC = log2FoldChange), by = "gene_symbol")

kable(comparison_table, digits = 3)
```

The hypoxia response was significantly reduced in MDA-MB-231 compared with MCF10A. This
is consistent with reports that triple negative breast cancer lines can already show a
partially hypoxic phenotype under normoxic conditions, meaning there is less room for
further induction.

The reduction was not the same for every gene. CA12, SPRY1 and ANKFN1 showed an almost
complete loss of hypoxia induction in MDA-MB-231, while EGLN3 (PHD3) remained
significantly induced in both cell lines, although more weakly in MDA-MB-231. This
suggests the reduced response may be specific to downstream target genes, while the
oxygen sensing machinery itself remains at least partly functional.

## 4.3 Hypoxia's transcriptional effect and the translating mRNA pool

```{r oxygen-fraction-interaction}
res_interaction_oxygen_fraction <- results(dds, name = "oxygenHypoxia.fractionPolysomal")
summary(res_interaction_oxygen_fraction)
```

No genes showed a significant oxygen:fraction interaction across the genome, even though
hypoxia affected both the Total and Polysomal fractions individually and to a similar
extent. This suggests the transcriptional response to hypoxia is passed through to the
polysome bound mRNA pool at a similar level, with no strong evidence of hypoxia specific
translational control in this dataset.

Limitation: with only 2 biological replicates per condition, this analysis does not have
enough power to detect smaller or gene specific interaction effects. This result should
be read as ruling out a large, genome wide effect, rather than ruling out translational
regulation of individual transcripts.

## 4.4 Pathway-level validation (GSEA)

```{r gsea}
gsea_input <- volcano_df %>%
  filter(!is.na(gene_symbol), !is.na(log2FoldChange)) %>%
  distinct(gene_symbol, .keep_all = TRUE)

gene_map <- bitr(gsea_input$gene_symbol, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

gsea_input <- gsea_input %>%
  inner_join(gene_map, by = c("gene_symbol" = "SYMBOL")) %>%
  distinct(ENTREZID, .keep_all = TRUE)

ranked_genes <- gsea_input$log2FoldChange
names(ranked_genes) <- gsea_input$ENTREZID
ranked_genes <- sort(ranked_genes, decreasing = TRUE)

hallmark_sets <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, entrez_gene)

set.seed(42)
gsea_result <- GSEA(geneList = ranked_genes, TERM2GENE = hallmark_sets, pvalueCutoff = 0.05, eps = 0)

gsea_df <- as.data.frame(gsea_result) %>% arrange(desc(NES))
kable(gsea_df[, c("ID", "NES", "pvalue", "p.adjust")], digits = 3, row.names = FALSE)

gseaplot2(gsea_result, geneSetID = "HALLMARK_HYPOXIA",
          title = "HALLMARK_HYPOXIA enrichment (Hypoxia vs Normoxia, MCF10A)")
```

HALLMARK_HYPOXIA was the most significantly enriched pathway, which supports the
differential expression results at the pathway level. The running enrichment score
(green line) shows the Hallmark Hypoxia genes concentrated among the most strongly
hypoxia upregulated transcripts in the ranked list, confirming the enrichment.

The wider results also formed a consistent pattern: HALLMARK_GLYCOLYSIS was upregulated
alongside hypoxia. This reflects the shift towards glycolytic metabolism that occurs
under low oxygen conditions. HALLMARK_OXIDATIVE_PHOSPHORYLATION was downregulated, which
fits with this shift, as cells rely less on oxidative phosphorylation when oxygen is
limited. Pathways linked to cell proliferation (HALLMARK_E2F_TARGETS,
HALLMARK_G2M_CHECKPOINT, HALLMARK_MYC_TARGETS_V1/V2) were also strongly suppressed. This
fits with the known hypoxia induced slowdown in cell cycle progression.

# 5. Discussion and Limitations

The aim of this analysis was to characterise the transcriptional response to hypoxia in
a non malignant breast cell line, and to determine whether this response differs in a
triple negative cancer cell line, or between total and polysome bound mRNA.

The results show a clear hypoxia signature in MCF10A cells. This is supported at
multiple levels: individual genes matched known HIF targets, the pathway level GSEA
result was strongly significant, and the PCA and heatmap both showed clean separation by
oxygen condition. MDA-MB-231 cells showed a much weaker version of this response, and
this reduction appeared to affect downstream effector genes more than the oxygen sensing
machinery itself.

A key strength of this analysis was the multi factor design, which made it possible to
test several biological questions from a single model rather than running separate
pairwise comparisons. This is a more statistically defensible approach given the limited
number of replicates available.

Several limitations should be considered:

- Replication: only 2 biological replicates were available per condition, which limits
  statistical power, particularly for interaction terms. The null oxygen:fraction result
  should not be taken as proof that no translational regulation occurs, only that a
  large effect was not detected.
- RSEM expected counts: rounding non integer expected counts to integers introduced minor
  artefacts in the dispersion estimates at low counts. These were corrected by DESeq2's
  shrinkage step.
- PP242 (drug) effect: the transcriptional effect of the mTOR inhibitor was comparatively
  small in this dataset and was not the focus of the main analysis. A more targeted
  analysis, for example restricted to known mTOR sensitive transcripts or the
  translational efficiency ratio, could investigate this further.

Overall, this analysis shows that a multi factor RNA seq design can be used to
characterise a known biological response (hypoxia) while also testing more specific
questions about how that response differs between cell types and mRNA fractions, within
a single reproducible pipeline.

# 6. References

- Korotkevich, G., Sukhov, V., Sergushichev, A. (2019) ‘Fast gene set enrichment analysis’. bioRxiv. doi:10.1101/060012

- Xu, S., Hu, E., Cai, Y., Xie, Z., Luo, X., Zhan, L., Tang, W., Wang, Q., Liu, B., Wang, R., Xie, W., Wu, T., Xie, L., Yu, G. (2024) ‘Using clusterProfiler to characterize multiomics data’. Nature Protocols, 19(11), pp.3292-3320. Doi: 10.1038/s41596-024-01020-z

- Zhu, A., Ibrahim, J.G., Love, M.I. (2018) ‘Heavy-tailed prior distributions for sequence count data: removing the noise and preserving large differences’. Bioinformatics, 35(12), pp.2084-2092. Doi:10.1093/bioinformatics/bty895

- Love, M.I., Huber, W. & Anders, S. (2014).'Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2'. Genome Biology, 15(550). doi:10.1186/s13059-014-0550-8

- Original dataset: GSE104193:
Sesé, M., Fuentes, P., Esteve-Codina, A., Béjar, E., McGrail, K., Thomas, G., Aasen, T., & Ramón Y Cajal, S. (2017). 'Hypoxia-mediated translational activation of ITGB3 in breast cancer cells enhances TGF-β signaling and malignant features in vitro and in vivo'. Oncotarget, 8(70), PP.114856–114876.DOI:10.18632/oncotarget.23145
